---
title: "Loading and Manipulating ACS Data in R"
output: 
  html_document:
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
ev <- FALSE
```

# Setup

Install the required packages by
```{r eval=FALSE}
install.packages(c("acs", "ggplot2", "openxlsx"))
```

For working with ACS data we will use the **acs** package which was developed by Ezra Haber Glenn (MIT), partly as a consulting project for the PSRC. Load into the namespace:
```{r eval=ev, message=FALSE, warning=FALSE}
library(acs)
```

Obtain Census API key from [here](http://api.census.gov/data/key_signup.html). Install the key on your system for all future sessions (replace the key below with your own key):

```{r eval=FALSE}
api.key.install("1596fcec45b7994ead5cd5d62a6542131df257f6")
```

# Loading ACS Data

First we need to determine for which geography and which dataset we want to obtain data.

* Lookup geography:

    ```{r eval=ev}
geo.lookup(state="WA", county="*")
geo.lookup(state="WA", county="Ki")
geo.lookup(state="WA", county="King", county.subdivision = "Seattle")
geo.lookup(state="WA", county="King", county.subdivision = "*")
```

* Create a geography:

    ```{r eval=ev}
# geography of all tracts in Kitsap
geo <- geo.make(state=53, county=35, tract="*", check = TRUE)
# geography of four counties
geo <- geo.make(state=53, county=c(33, 35, 53, 61), check=TRUE)
```

* Lookup ACS table. Say, we are looking for a table of the ratio of income to poverty level:
    
    ```{r eval=ev}
acs.lookup(2015, span = 1, table.name = c("Income", "Poverty"), case.sensitive = FALSE)
acs.lookup(2015, span = 1, table.number = "B17002")
```

* Loading data:

    ```{r eval=ev}
dataset <- acs.fetch(endyear = 2015, span = 1, geography = geo, table.number = "B17002") 
dim(dataset)
dataset
dataset <- acs.fetch(endyear = 2015, span = 1, geography = geo, table.number = "B17002", 
                     col.names="pretty") 
dataset
```

* Change column names. Remove the table name from column names:

    ```{r eval=ev}
acs.colnames(dataset)
acs.colnames(dataset) <- gsub(
  "Ratio of Income to Poverty Level in the Past 12 Months: ", "", 
  acs.colnames(dataset))
dataset
```

# Manipulating and Visualizing ACS Data

* Various mathematic operations are defined on acs datasets. For example, to create a dataset of shares, do:

    ```{r eval=ev}
shares <- divide.acs(dataset, dataset[,rep(1, ncol(dataset))])*100
acs.colnames(shares) <- acs.colnames(dataset)
```

* Sums over columns, rows or individual cells:

    ```{r eval=ev}
apply(dataset, 1, sum)
apply(dataset, 2, sum)
dataset[,2] + dataset[,3] + dataset[,4]
dataset[2,2] + dataset[3,2]
```

* Visualization:

    ```{r eval=ev}
plot(shares)
plot(shares[,-1])
plot(dataset[,2:7], by = "acs.colnames", labels = c("King", "Kitsap", "Pierce", "Sno"))
plot(sum(dataset[,1]))
```

* Adding rows/columns. Add a row of sums:

    ```{r eval=ev}
data.wsum <- rbind(dataset, apply(dataset, 1, sum))
```

* Converting to R matrices 

    Decompose into estimates and standard errors:
    
    ```{r eval=ev}
est <- estimate(data.wsum)
est
class(est)
se <- standard.error(data.wsum)
se
rownames(est) <- rownames(se) <- gsub(" County, Washington", "", rownames(est))
shest <- estimate(shares)
shse <- standard.error(shares)
rownames(shest) <- rownames(shse) <- rownames(est)[1:4]
```

* View with standard R graphics:

    ```{r eval=ev}
plot(est[,1], se[,1])
plot(est[2,-1], est[3,-1])
hist(est[5,-1])
```

* Fancy plots via `ggplot2`
    * by variable:
    
        ```{r eval=ev}
library(ggplot2)
library(reshape2)
df <- melt(shest[,-1])
head(df)
colnames(df)[1:2] <- c("County", "Variable")
g1 <- ggplot(df, aes(Variable, fill=County, weight=value)) + 
  geom_bar(position="dodge") + ylab("") + xlab("")
print(g1)
```

    * by county:
    
        ```{r eval=ev}
g2 <- ggplot(df, aes(County, fill=Variable, weight=value)) + 
  geom_bar(position="dodge") + ylab("") + xlab("")
print(g2)
```

* Save into a file
    * Estimates only:
    
        ```{r eval=FALSE}
write.csv(est, file="population_by_income.csv")
write.csv(t(est), file="population_by_income_transposed.csv")
```

    * Including 95% confidence intervals:
<!--90%: qnorm(1-(1-0.90)/2)-->

        ```{r eval=FALSE}
out <- matrix(NA, ncol=ncol(est)*2, nrow=nrow(est))
colidx <- seq(1, ncol(out), by=2)
out[, colidx] <- est
out[, colidx+1] <- round(1.96*se, 2)
rownames(out) <- rownames(est)
coln <- rep(colnames(est), each=2)
coln[colidx+1] <- paste(coln[colidx+1], "(MoE)")
colnames(out) <- coln
write.csv(out, file="population_by_income_inclMoE.csv")
```


# Functions
* Write a function that combines estimates and margin of errors (store the code below into a new file):

    ```{r eval=ev}
create.combined.matrix <- function(est, moe, decimals=2) {
    out <- matrix(NA, ncol=ncol(est)*2, nrow=nrow(est))
    colidx <- seq(1, ncol(out), by=2)
    out[, colidx] <- est
    out[, colidx+1] <- round(moe, decimals)
    rownames(out) <- rownames(est)
    coln <- rep(colnames(est), each=2)
    coln[colidx+1] <- paste(coln[colidx+1], "(MoE)")
    colnames(out) <- coln
    return(out)
}
```

    Source the file and call the function with:
   
    ```{r eval=ev}  
create.combined.matrix(est, 1.64*se)    
```

* Create a function to run the above steps for different tables and different years. 
    * Store the code below into the same file as ``create.combined.matrix`` (it is a collection of steps from above):

        ```{r eval=ev}
popbyincome <- function(table, year = 2015) {
    # create geography and load data
    geo <- geo.make(state = 53, county = c(33, 35, 53, 61))
    dataset <- acs.fetch(endyear = year, span = 1, geography = geo, table.number = table)
    # shorten column names
    acs.colnames(dataset) <- gsub(
        "Ratio of Income to Poverty Level in the Past 12 Months: ", "", 
        acs.colnames(dataset))
    # add row of sums
    data.wsum <- rbind(dataset, apply(dataset, 1, sum))
    # extract estimates & MoE
    est <- estimate(data.wsum)
    se <- standard.error(data.wsum)
    # shorten row names
    rownames(est) <- rownames(se) <- gsub(" County, Washington", "", rownames(est))
    # combine & write into file
    result <- create.combined.matrix(est, 1.96*se)
    write.csv(result, file=paste0("population_by_income_", table, ".csv"))
}
```

    * Create a loop that calls the above function with different values of the arguments:
    
        ```{r eval=FALSE}
for(year in c(2014, 2015)) {
    for (tab in c('B17002', 'B17022', 'B17024')) {
        popbyincome(tab, year=year)
    }
}
```

# Working with Other Data Formats

* Reading ASCII files:
    * CSV:
    
        ```{r eval=ev}
df <- read.csv("population_by_income.csv")
head(df)
df <- read.csv("population_by_income.csv", check.names = FALSE)
head(df)
```

    * More generic:

        ```{r eval=ev}
df <- read.table("population_by_income.csv", check.names = FALSE, 
                  sep = ",", header = TRUE)
head(df)
```

    * File converter:
  
        ```{r eval=ev}
convert <- function(in.file, out.file, from, to, ...) {
  df <- read.table(in.file, sep=from, ...)
  write.table(df, file=out.file, sep=to)
}

convert("population_by_income.csv", "population_by_income.txt", from=",", to="\t", 
          header = TRUE, check.names = FALSE)
```
  
* Downloading online datasets. For example, download population dataset from the OFM site http://www.ofm.wa.gov:

    ```{r eval=FALSE}
download.file("http://www.ofm.wa.gov/pop/census2010/pl/data/wa_2010_pl94171_city.zip", 
                "wa2010city.zip")
file.name <- unzip("wa2010city.zip")
```

* Reading Excel files:

    ```{r eval=ev}
library(openxlsx)
getSheetNames(file.name)
df <- read.xlsx(file.name, sheet="Table1")
View(df)
df <- read.xlsx(file.name, sheet="Table1", startRow = 4)
head(df)
# make the first column to be rownames
rownames(df) <- df[,1]
df <- df[,-1]
```

* Exporting into Excel files:

    ```{r eval=ev}
write.xlsx(df, "mydataset.xlsx", sheetName="mysheet")
write.xlsx(t(df), "mydataset.xlsx", sheetName="transposed", append=TRUE)
```

# Summary of Common Data Manipulation Tasks

Let ``df`` be an object of class ``data.frame``.

  * Change column names of a data frame: ``colnames(df) <- c(...)``
  * Change row names of a data frame: ``rownames(df) <- c(...)``
  * Extract column: 
    * ``df$my_column``
    * ``df[,"my_column"]``
    * ``df[,colindex]``
    * ``df[, which(colnames(df) == "my_column")]``
  * Extract row:
    * ``df["my_row"]``
    * ``df[rowindex,]``
    * ``df[which(rownames(df) == "my_row")]``
  * Add column: ``cbind(df, new_column=values)``
  * Add row: ``rbind(df, values_or_data_frame)``
  * Remove column:
    * ``df$my_columns <- NULL``
    * ``df[,-colindex]``
    * ``df[, -which(colnames(df) == "my_column")]``
  * Remove row:
    * ``df[-rowindex,]``
    * ``df[-which(rownames(df) == "my_row")]``

# Useful Links
  * [2016 Software Carpentry R Workshop](http://psrc.github.io/novice-r)
  * [acs-r mailing list](http://mailman.mit.edu/mailman/listinfo/acs-r)
  * [example of creating a map with acs data](http://eglenn.scripts.mit.edu/citystate/2013/07/using-acs-r-to-create-chorpleth-maps)
  

